Evaluating Deep Learning Methods for Tokenization of Space-less texts in Old French

### 2019-06-04 article/article.tex
Updated of the article

\abstract{Tokenization of modern and old Western European languages seems to be fairly simple as it stands on the presence mostly of markers such as spaces and punctuation. Although, when dealing with old sources like manuscript written in \textit{scripta continua}, (1) such markers are mostly absent, (2) spelling variation and rich morphology makes dictionary based approaches difficult. We show that applying convolutional encoding to characters followed by linear categorization to word-boundary or in-word-sequence can be used to tokenize such inputs. Additionally, we release a software with a rather simple interface for tokenizing one's corpus.}

@@

The training parameters was 0.00005 in learning rate for each CNN model and 0.001 for the LSTM one, and 64 in batch sizes. Training reached a plateau fairly quickly for each model (\textit{cf.} \ref{fig:loss}). Each model except LSTM reached a really low loss and a high accuracy on the test set (\textit{cf.} \ref{tab:scores})

@@

\begin{table}[!ht]
\centering
\begin{tabular}{lllll}
\hline
Model   & Accuracy & Precision & Recall & FScore \\ \hline
CNN     & 0.991    & 0.985     & 0.990  & 0.987  \\
CNN L   & 0.991    & 0.979     & 0.990  & 0.985  \\
CNN P   & \textbf{0.993}    & \textbf{0.990}& \textbf{0.991}  & \textbf{0.990}  \\
CNN N   & 0.991    & 0.987     & 0.988  & 0.988  \\
CNN L N & 0.992    & 0.988     & 0.989  & 0.988  \\
LSTM    & 0.741    & 0.184     & 0.500  & 0.269  \\ \hline
\end{tabular}
\caption{Scores over the test dataset. N = normalized, L = Lower, P = no position embedding.}
\label{tab:scores}
\end{table}

@@

>>>>

\abstract{Tokenization of modern and old Western European languages seems to be fairly simple as it stands on the presence mostly of markers such as spaces and punctuation. Although, when dealing with old sources like manuscript written in \textit{scripta continua} or later manuscripts, (1) such markers are mostly absent, (2) spelling variation and rich morphology makes dictionary based approaches difficult. We show that applying convolutional encoding to characters followed by linear categorization to word-boundary or in-word-sequence can be used to tokenize such inputs. Additionally, we release a software with a rather simple interface for tokenizing one's corpus.}

@@

The training parameters was 0.00005 in learning rate for each CNN model and 0.001 for the LSTM one, and 64 in batch sizes. Training reached a plateau fairly quickly for each model (\textit{cf.} \ref{fig:loss}). Each model except LSTM reached a really low loss and a high accuracy on the test set (\textit{cf.} \ref{tab:scores}). To compare the results, we used the \texttt{wordsegment} package \citet{WordSegment} as a baseline.

@@

\begin{table}[!ht]
\centering
\begin{tabular}{lllll}
\hline
Model    & Accuracy & Precision & Recall & FScore \\ \hline
Baseline & 0.989    & 0.986     & 0.984  & 0.985 \\
CNN      & 0.991    & 0.985     & 0.990  & 0.987  \\
CNN L    & 0.991    & 0.979     & 0.990  & 0.985  \\
CNN P    & \textbf{0.993}    & \textbf{0.990}& \textbf{0.991}  & \textbf{0.990}  \\
CNN N    & 0.991    & 0.987     & 0.988  & 0.988  \\
CNN L N  & 0.992    & 0.988     & 0.989  & 0.988  \\
LSTM     & 0.741    & 0.184     & 0.500  & 0.269  \\ \hline
\end{tabular}
\caption{Scores over the test dataset. N = normalized, L = Lower, P = no position embedding.}
\label{tab:scores}
\end{table}

\subsubsection{Unknown texts}

While all model using CNN shows improvement over the baseline, the model definitely does not outperform it by a huge margin (\textless 0.02 FScore). And for a reason : the baseline already performs nearly perfectly on the test corpus. The dictionary attack using n-grams did actually perform well. As a result, we wanted to compare how both models would perform on a secondary test corpus composed by texts that were not used in training: indeed, the training, dev and test corpus share the same texts while not sharing the same inputs. As a result, we created a new corpus with 4 texts and 742 samples : the diplomatic edition of the Graal \citet{graal}, a Passion and a Vie de Saint Leger \citet{old_french_corpus}, a Vie de Saint Thibaut \citet{theobaldus}. No noise characters and no random keeping of words were applied.

The results here were highly different (\textit{cf.} Table \ref{tab:scores_unknown}): while it appears that the CNN is able to expand its "comprehension" of the language to newer texts, the new words are more difficult to take into account for the baseline \texttt{wordsegment} n-gram approach, resulting in a respective drop to 0.945 and 0.838 FScore. WordSegment specifically performed badly with WB false positives : it had 3658 over a corpus containing 10,193 WB token (around 35 \%).

\begin{table}[!ht]
\centering
\begin{tabular}{lllllll}
\hline
 & Accuracy & Precision & Recall & FScore & WB as WC & WC as WB \\ \hline
Baseline & 0.882 & 0.893 & 0.808 & 0.838 & 3658 & 644 \\
CNN P & \textbf{0.957} & \textbf{0.948} & \textbf{0.944} & \textbf{0.945} & \textbf{854} & \textbf{723} \\ \hline
\end{tabular}
\caption{Scores over the unknown dataset. WB as WC and WC as WB are taken from the confusion matrices}
\label{tab:scores_unknown}
\end{table}

@@

While the baseline surprised us by performing this well on the test corpus, it definitely performed less well than the CNN on a completely unknown corpus: in this context, the proposed model actually shows its ability to carry over unknown corpora in a better way than classical ngram approaches.
